---
title: "Bayesian Boundaries: Investigating Credible Intervals (Context-Free)"
format: html
theme: flatly
draft: true
toc: true
date: "09/13/2024"
date-modified: today
date-format: long
bibliography: act2bib.bibtex
author: 
  - name: "Adam Gilbert"
    affiliations: 
      - name: "Southern New Hampshire University"
  - name: "Laura Lambert"
    affiliations: 
      - name: "James Madison University"
engine: knitr
webr:
  channel-type: 'automatic'
  packages: ["ggplot2", "dplyr", "bayesrules"]
  
filters:
  - webr
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(kable_styling_bootstrap_options = c("hover", "striped"))

library(tidyverse)
library(kableExtra)
library(bayesrules)
library(patchwork)

theme_set(theme_bw(base_size = 14))


##Need to deal with data "read in" via webr cells not available at render time.
##This means inline R with that data and associated values is not possible.
##For now, I'm just adding in the values to make this work -- I'll likely 
##pre-read data in, compute the necessary values, and then remove the data frame
new_obs <- 157
new_pos <- 102
```

:::{.callout-note}
## Context-Free Activity
This activity is a *context-free scaffold*. You'll almost surely find it to be incomplete, and that's on purpose! We are designing this series of activities to have a *domain-specific* context laid over them. Throughout the activity, you'll see purple text, which is text that should be replaced with context-specific content. A rough draft completed activity is [available here, in the chytrid fungus context](https://agmath.github.io/Bayes-BATS-tier2/act2_bio_chytrid.html).
:::

## Goals and Objectives

**Statistics Goals:** The statistics-focused goals of this activity are as follows:

+ Recall the concepts of *prior* and *posterior* distributions and use the `{bayesrules}` package to easily update a *prior* with observed data, obtaining a *posterior* distribution.
+ Extract a credible interval from a *posterior* distribution
+ Interpret a credible interval obtained from a *posterior* distribution

**Course Objectives:**  This activity would map to course-level objectives similar to the following.  Note that this is not an exhaustive list, nor should it be interpreted that objectives must be phrased identically.  This is to give an idea as to the wide variety of contexts this activity might be placed in.

+ Students will evaluate a research question using appropriate statistical techniques
+ Students will correctly identify the type of data they are working with
+ Students will evaluate literature and/or prior research to generate hypotheses for a research question
+ Students will learn about different statistical models and approaches
+ Students will interpret coefficients from a statistical model
+ Students will evaluate the underlying assumptions of a statistical approach
+ Students will consider the ethical implications of statistical approaches
+ Students will gather data using methodologies appropriate to the context

<font color="purple">**Subject-Specific Objectives:** This section is where course-specific objectives relevant to the activity’s context can be detailed. This allows instructors to meet objectives specific to their discipline while incorporating Bayesian thinking.</font>

## Background Information

The following sections provide the statistical and contextual background for this activity.

### Data Analysis and Bayesian Thinking

There are many statistical tools which can be used to investigate population parameters. As we mentioned in [our first activity](https://agmath.github.io/Bayes-BATS-tier2/ContextFreeActivity1.html), these tools fall into three broad categories:

+ Classical/Frequentist methods 
+ Simulation-based methods
+ Bayesian methods

This series of activities focuses on Bayesian methods. In this second activity, we engage in extracting and interpreting a *credible interval* from a *posterior* distribution. We’ll briefly review how to obtain the *posterior* distribution from the *prior* and observed data, but check back on [our first activity](https://agmath.github.io/Bayes-BATS-tier2/ContextFreeActivity1.html) for full details.

### About the Context

<font color="purple">This subsection includes background on the domain-specific context for the activity.</font>

### Purpose

Building on concepts from the first activity, we'll continue to estimate the <font color="purple">[relevant population parameter]</font> by extracting and interpreting a *credible interval* from the *posterior distribution*.

## Reminder About Prior and Posterior Distributions

In Bayesian inference, a *prior distribution* represents initial beliefs about a population parameter. This allows for prior knowledge or experiences (and uncertainties) to inform conclusions without solely relying on new data. The *posterior distribution* is then obtained by updating the *prior* with observed data by using a version of Bayes' Rule:

$$\mathbb{P}\left[\text{parameter value} \mid \text{data}\right] = \frac{\mathbb{P}\left[\text{data} \mid \text{parameter value}\right]\cdot \mathbb{P}\left[\text{parameter value}\right]}{\mathbb{P}\left[\text{data}\right]}$$

In [Activity 1](https://agmath.github.io/Bayes-BATS-tier2/ContextFreeActivity1.html), we manually updated the prior, applying this relationship step-by-step via code, which provided transparency into the Bayesian updating process.

### Convenient Functionality from `{bayesrules}`

In this activity, we’ll simplify the process of obtaining our *posterior* distributions by using functions from the `{bayesrules}` package. *Bayes Rules! An Introduction to Bayesian Modeling*, by Johnson, Ott, and Dogucu [@bayesrulesbook] and the corresponding `{bayesrules}` R package [@bayesrulespkg] provide user-friendly functionality that we’ll take advantage of.

In our first activity, we began with an *uninformative prior*, represented as <font color="purple">a beta distribution with one observed success and one failure, `beta(1, 1)`</font>. We then updated that prior with our <font color="purple">initially observed data including eight (8) successes out of twelve (12) observations</font>. Using the `plot_beta_binomial()` and `summarize_beta_binomial()` functions from `{bayesrules}`, we can easily visualize and summarize the prior, likelihood, posterior, and associated statistics.

To use `plot_beta_binomial()`, we need to specify four arguments: 

+ `alpha` is the number of previously observed successes
+ `beta` is the number of previously observed failures
+ `y` is the number of newly observed successful outcomes
+ `n` is the total number of new observations recorded

Below is an example with our initial `beta(1, 1)` *prior* and the first set of new observations we encountered:

```{r}
plot_beta_binomial(alpha = 1, beta = 1, y = 8, n = 12)
```

In the plot, the prior distribution appears in yellow, while the posterior is shown in blue-green. Here, our uninformative prior results in the posterior overlapping exactly with the scaled likelihood. This is because of our choice to use the uniform prior.

To obtain summary statistics on the prior and posterior, we use the `summarize_beta_binomial()` function, which provides key characteristics of each distribution, including the *mean*, *mode*, *standard deviation*, and *variance*. This function takes the same arguments as `plot_beta_binomial()`.

```{r}
#| echo: true
#| eval: false

summarize_beta_binomial(alpha = 1, beta = 1, y = 8, n = 12)
```

```{r}
#| echo: false
#| eval: true

summarize_beta_binomial(alpha = 1, beta = 1, y = 8, n = 12) %>%
  kable() %>%
  kable_styling()
```

We'll quickly revisit the second investigation we made in that initial activity as well. <font color="purple">You may remember that we obtained advice from an expert, indicating that we suspect that the population proportion should be near 60%, given prior research results</font>. At that point, we re-evaluated our choice of *prior* distribution to reflect this. <font color="purple">We chose a *weak prior*, which was still a beta distribution but with 6 previously observed successes and 4 previously observed failures. Such a prior is *weak* because it assumes information from very few previous observations.</font>

Let's plot that prior, along with the scaled likelihood and posterior distribution using `plot_beta_binomial()`. We'll also obtain the numerical summaries for the prior and posterior distributions while we are at it.

```{r}
plot_beta_binomial(alpha = 6, beta = 4, y = 8, n = 12)
```

```{r}
#| echo: true
#| eval: false

summarize_beta_binomial(alpha = 6, beta = 4, y = 8, n = 12)
```

```{r}
#| echo: false
#| eval: true

summarize_beta_binomial(alpha = 6, beta = 4, y = 8, n = 12) %>%
  kable() %>%
  kable_styling()
```

From the output above, we can see that the *prior*, the scaled likelihood, and the *posterior* are all distinct. The *prior* and the *likelihood* of our observed data act together to arrive at the *posterior* distribution. Again, the use of `plot_beta_binomial()` simplifies our job by performing the Bayesian update behind the scenes.

While we are transitioning to the more "black box" functionality in this activity (and you'll continue with similar functionality in the remaining activities too), the foundation you built in engaging with the first activity should give you some intuition about what functions like `plot_beta_binomial()` and `summarize_beta_binomial()` are doing when they are run. 

:::{.callout-caution}
## Discussion Question

Take this opportunity to discuss what is happening "behind the scenes" when the `plot_beta_binomial()` function is run.
:::

## Parameter Estimation with Intervals

Through the main section of this activity, you'll explore the use of *credible intervals* (CIs), the Bayesian analog to the Frequentist's *confidence interval*. A key advantage of credible intervals is that they offer a much more intuitive interpretation. For example, with a 95% credible interval, we claim there is a 95% probability that the population parameter lies within the interval bounds.

Before we get into CIs, let's get ourselves back to where we were at the end of the last activity. <font color="purple">As you might remember, we obtained the following data. 

:::{.callout-important} 
## About Data

As a reminder, we can set up this code chunk so that the data is

+ simulated, as in the code chunk below.
+ read in from a location on the web (for example, a GitHub repository). This could be data from a study, a publication, your own research, etc.
+ collected in class and manually input into the `status` column of the *tibble* below. 
  
  + This is an unlikely choice due to the relatively small sample size and time required to input data. This method could be accommodated if data is collected via a digital form and then read from a spreadsheet (ie. Google Forms/Sheets).

:::

The code below will be replaced based on the choice made above.
</font>

```{webr-r}
new_obs <- 157
set.seed(072324)
my_data <- tibble(
  obs_number = 1:new_obs,
  status = sample(c("yes", "no"), 
                  size = new_obs,
                  prob = c(0.7, 0.3), 
                  replace = TRUE)
)

new_pos <- my_data %>%
  filter(status == "yes") %>%
  nrow()

print(paste0("Of the ", new_obs, 
             " observations, there were ", 
             new_pos, " positive responses."))
```

Use the `plot_beta_binomial()` and `summarize_beta_binomial()` functions in the code cell below to update your <font color="purple">weakly informative `beta(6, 4)` prior with these `r new_obs` observations including `r new_pos` positive outcomes.</font>

```{webr-r}

```

:::{.callout-caution}
## Discussion Question

When observing the results from these two functions, consider the following questions.

i) Visually, what do you notice about the most likely values for the population parameter?
ii) Numerically, what insights can we draw from the summary statistics we've obtained about the prior and posterior distributions?
:::

Notice that this new data updates our prior beliefs about our population parameter. The posterior distribution now represents our updated "world-view" after seeing new evidence.

### Credible Intervals

If you’ve taken a traditional (Frequentist) statistics course, you might be familiar with *confidence intervals*. These intervals aim to capture a population parameter with a prescribed level of certainty. In practice, however, interpreting these intervals can be tricky. Confidence intervals are based on a hypothetical scenario in which we take many random samples and build intervals from each; the expectation is that, over many intervals, a certain proportion (like 95% for a 95% confidence interval) would contain the true parameter.

In Bayesian inference, *credible intervals* offer a simpler, probabilistic interpretation. A 90% credible interval extends from the 5th percentile to the 95th percentile of the posterior distribution. This means there’s a 90% probability that the population parameter falls within these bounds, based on the prior assumption and the observed data — no hypothetical repetitions required.

In this section of our activity, we'll compute and interpret credible intervals. To do so, we'll need a bit of information about the `qbeta()` function, which calculates "quantiles" from a beta distribution. To use `qbeta()`, we input the proportion of observations to the left of the desired quantile, along with the $\alpha$ and $\beta$ parameters defining the distribution. This returns the cutoff value at the desired quantile.

<font color="purple">For example, we calculate the $5^{\text{th}}$ and $95^{\text{th}}$ percentiles from our weakly informative `beta(6, 4)` prior, giving us a 90% credible interval below.

```{r}
lower_bound <- qbeta(0.05, 6, 4)
upper_bound <- qbeta(0.95, 6, 4)

lower_bound
upper_bound
```

So, based on our prior, we have a 90% probability that the population parameter falls between about 34.5% and about 83.1%. Note that we could have also calculated both of these bounds at once using `qbeta(c(0.05, 0.95), 6, 4)`.

We can also visualize this confidence interval by using the `plot_beta_ci()` function, which takes $\alpha$, $\beta$ and the desired level of coverage as arguments.

```{r}
plot_beta_ci(6, 4, 0.90)
```

</font>


Now, use the updated data (<font color="purple">the `r new_pos` positive outcomes from the `r new_obs` observations</font>) and your results from `summarize_beta_binomial()`, along with the `qbeta()` function to extract a new 90% credible interval for the population proportion from your *posterior* distribution. Use `plot_beta_ci()` to plot the credible interval as well.

```{webr-r}

```

:::{.callout-caution}
## Discussion Questions

i) How does the credible interval extracted from the *posterior* distribution differ from the one extracted from the *prior* distribution? What does that mean? 
ii) What does this new credible interval tell you about the population parameter? Discuss what these specific interval bounds mean in practical terms for the population parameter being investigated. 
iii) How might this interval change if we collected more data or desired greater coverage. Under each scenario, would the interval be more narrow or become wider? Why?
:::

Now calculate a 95% credible interval for the population proportion. 

```{webr-r}

```

:::{.callout-caution}
## Discussion Questions

i) Compare your 90% and 95% credible intervals. Why does the 95% credible interval contain a wider range of values? Did this match your expectation from above?
ii) What do we gain by using a wider interval? Are there practical contexts in which you might prefer a narrower (e.g., 90%) credible interval over a wider one, or vice versa?
:::

## Summary

In this notebook we,

+ reinforced the roles of the *prior* and *posterior* distributions.
+ saw that our observed data updates our *prior* beliefs, resulting in a new understanding of the world we are investigating.
+ constructed and utilized *credible intervals* to capture a population parameter.
+ we interpreted our *credible intervals* in a meaningful context and by using the probablistic and intuitive language that the Bayesian approach affords us.

:::{.callout-caution}
## Reflection Question

Reflect on where you could apply this new knowledge of credible intervals. Consider examples from:

i) **Your current coursework:** How might credible intervals support your analyses in other courses?
ii) **Personal interests:** Are there topics or hobbies where interval estimates could clarify uncertainty?
iii) **Professional projects:** Think about how credible intervals might apply in work settings where you estimate uncertain quantities.
:::

## References

:::{.refs}

:::