---
title: "Bayesian Boundaries: Investigating Credible Intervals (Context-Free)"
format: html
theme: flatly
draft: true
toc: true
date: "09/13/2024"
date-modified: today
date-format: long
bibliography: act2bib.bibtex
author: 
  - name: "Adam Gilbert"
    affiliations: 
      - name: "Southern New Hampshire University"
  - name: "Laura Lambert"
    affiliations: 
      - name: "James Madison University"
engine: knitr
webr:
  channel-type: 'automatic'
  packages: ["ggplot2", "dplyr", "bayesrules"]
  
filters:
  - webr
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(kable_styling_bootstrap_options = c("hover", "striped"))

library(tidyverse)
library(kableExtra)
library(bayesrules)

theme_set(theme_bw(base_size = 14))


##Need to deal with data "read in" via webr cells not available at render time.
##This means inline R with that data and associated values is not possible.
##For now, I'm just adding in the values to make this work -- I'll likely 
##pre-read data in, compute the necessary values, and then remove the data frame
new_obs <- 157
new_pos <- 102
```

:::{.callout-note}
## Context-Free Activity
This activity is a *context-free scaffold*. You'll almost surely find it to be incomplete, and that's on purpose! We are designing this series of activities to have a *domain-specific* context laid over them. Throughout the activity, you'll see purple text, which is text that should be replaced with context-specific content. A rough draft completed activity will be [made available here when it is ready](https://agmath.github.io/Bayes-BATS-tier2/act2_bio_chytrid.html).
:::

## Goals and Objectives

**Statistics Goals:** The statistics-focused goals of this activity are as follows:

+ Recall the notion of the *prior* and *posterior* distributions
+ Extract a credible interval from a *posterior* distribution
+ Interpret a credible interval obtained from a *posterior* distribution

**Course Objectives:**  This activity would map to course-level objectives similar to the following.  Note that this is not an exhaustive list, nor should it be interpreted that objectives must be phrased identically.  This is to give an idea as to the wide variety of contexts this activity might be placed in.

+ Students will evaluate a research question using appropriate statistical techniques
+ Students will correctly identify the type of data they are working with
+ Students will evaluate literature and/or prior research to generate hypotheses for a research question
+ Students will learn about different statistical models and approaches
+ Students will interpret coefficients from a statistical model
+ Students will evaluate the underlying assumptions of a statistical approach
+ Students will consider the ethical implications of statistical approaches
+ Students will gather data using methodologies appropriate to the context

<font color="purple">**Subject-Area Objectives:** This section will be utilized to identify objectives/outcomes specific to the course/subject to which the activity context is linked. This allows adopters to cover objectives associated with their course while embedding Bayesian thinking.</font>

## Background Information

The following subsections outline the background information for this activity from both a statistics and domain-specific lens. 

### Data Analysis and Bayesian Thinking

There are many statistical tools which can be used to test hypotheses about population parameters. Broadly speaking, these tools fall into three categories:

+ Classical/Frequentist methods 
+ Simulation-based methods
+ Bayesian methods

Perhaps you've encountered frequentist methods previously. These methods depend on distribution assumptions and the Central Limit Theorem. In this notebook, we engage in extracting and interpreting a credible interval from a *posterior* distribution. We'll provide a quick recap on the *posterior* distribution and how it is obtained from our *prior* distribution and out observed data in this notebook, but check back on [our first activity](https://agmath.github.io/Bayes-BATS-tier2/ContextFreeActivity1.html) for more details.

### About the Context

<font color="purple">This subsection includes background on the domain-specific context for the activity.</font>

### Purpose

Let's build off of what we learned in the first activity and continue trying to estimate the <font color="purple">[population parameter in context]</font> by extracting a *credible interval* from the *posterior distribution*.

## Reminder About Prior and Posterior Distributions

In Bayesian Inference, a *prior distribution* is an expression of our prior beliefs about a population parameter. It acknowledges our uncertainties while, at the same time, allowing prior knowledge or experiences to have some influence on our conclusions. In this way, we aren't depending on the randomly sampled data to completely determine our findings.

In [our first activity](https://agmath.github.io/Bayes-BATS-tier2/ContextFreeActivity1.html), we saw that we can obtain the *posterior* distribution from our *prior* and our data by making use of a form of Bayes' Rule.

$$\mathbb{P}\left[\text{parameter value} \mid \text{data}\right] = \frac{\mathbb{P}\left[\text{data} \mid \text{parameter value}\right]\cdot \mathbb{P}\left[\text{parameter value}\right]}{\mathbb{P}\left[\text{data}\right]}$$

In that activity, we did all of the Bayesian updating to obtain the prior "by hand". That is, we built the *posterior* distribution by explicitly applying the relationship above via code. You could read through the code and see exactly how and where the update was being made.

### Convenient Functionality from `{bayesrules}`

Beginning with this activity, we'll shift to less transparent, but more convenient, functionality to conduct the Bayesian updating from here on out. The *Bayes Rules! An Introduction to Bayesian Modeling* book written by Johnson, Ott, and Dogucu [@bayesrulesbook] along with their corresponding `{bayesrules}` R package [@bayesrulespkg] provides the excellent functionality that we'll benefit from.

If you recall, in that first activity we had begun by assuming a completely uninformative prior -- <font color="purple">a beta distribution with one observed success and one observed failure</font>. We then updated that prior with our <font color="purple">initially observed data including eight (8) successes out of twelve (12) observations</font>. We can quickly obtain a plot of our *prior*, our data *likelihood*, our *posterior*, and some summary statistics just by using the `plot_beta_binomial()` and `summarize_beta_binomial()` functions from `{bayesrules}`.

**Add paragraph with information about parameters for `plot_beta_binomial()` as well as the results from running that function -- what do the results "mean" and how can we utilize them?**

```{r}
plot_beta_binomial(alpha = 1, beta = 1, y = 8, n = 12)
```

From the plot above, we can see our assumed *prior* distribution in yellow and our *posterior* distribution in a purplish color. Note that the *posterior* and the scaled *likelihood* overlap exactly here because of our assumed uniform prior.

Now we'll use `summarize_beta_binomial()` to obtain summary statistics on the *prior* and *posterior* distributions. **Add discussion about parameters for `plot_beta_binomial()` as well as the results from running that function -- what do the results "mean" and how can we utilize them?**

```{r}
#| echo: true
#| eval: false

summarize_beta_binomial(alpha = 1, beta = 1, y = 8, n = 12)
```

```{r}
#| echo: false
#| eval: true

summarize_beta_binomial(alpha = 1, beta = 1, y = 8, n = 12) %>%
  kable() %>%
  kable_styling()
```

We'll quickly revisit the second thing we did in that activity as well. You may remember that we got advice from an expert, indicating that <font color="purple">we suspect that the population proportion should be near 60%, given prior research results</font>. At that point, we re-evaluated our choice of *prior* distribution to reflect this. <font color="purple">We chose a *weak prior*, which was still a beta distribution but with 6 previously observed successes and 4 previously observed failures. Such a prior is weak because it assumes information from very few previous observations.</font>

Let's plot that prior, along with the scaled likelihood and posterior distribution using `plot_beta_binomial()` and also obtain the numerical summaries for the prior and posterior distributions while we are at it.

```{r}
plot_beta_binomial(alpha = 6, beta = 4, y = 8, n = 12)
```

```{r}
#| echo: true
#| eval: false

summarize_beta_binomial(alpha = 6, beta = 4, y = 8, n = 12)
```

```{r}
#| echo: false
#| eval: true

summarize_beta_binomial(alpha = 6, beta = 4, y = 8, n = 12) %>%
  kable() %>%
  kable_styling()
```

From the output above, we can see that the *prior*, the scaled likelihood, and the *posterior* are all distinct. The *prior* and the *likelihood* of our observed data act together to arrive at the *posterior* distribution.

While we are transitioning to the more "black box" functionality in this activity (and you'll continue with similar functionality in the remaining activities too), the foundation you built in engaging with the first activity should give you some intuition about what functions like `plot_beta_binomial()` and `summarize_beta_binomial()` are doing "under the hood". 

:::{.callout-caution}
## Reflection Question

Take this opportunity to discuss what is happening "behind the scenes" when the `plot_beta_binomial()` function is run.
:::

## Parameter Estimation with Intervals

Through the main section of this activity, you'll explore two types of intervals which can be used to estimate the location of a population parameter. The first is the *credible interval* (CIs), which is the Bayesian analog to the Frequentist's *confidence interval*. The second is the *highest density probability interval* (HDPIs), which may do a better job than the credible interval does at identifying the "most likely" location of the population parameter.

Before we get into CIs and HDPIs, let's get ourselves back to where we were at the end of the last activity. <font color="purple">As you might remember, we obtained the following data. 

As a reminder, we can set up this code chunk so that the data is

+ simulated, as in the code chunk below.
+ read in from a location on the web (for example, a GitHub repository). This could be data from a study, a publication, your own research, etc.
+ collected in class and manually input into the `status` column of the *tibble* below. 
  
  + This is an unlikely choice due to the relatively small sample size and time required to input data. This method could be accommodated if data is collected via a digital form and then read from a spreadsheet (ie. Google Forms/Sheets).

The code below will be replaced based on the choice made above.
</font>

```{webr-r}
new_obs <- 157
set.seed(072324)
my_data <- tibble(
  obs_number = 1:new_obs,
  status = sample(c("yes", "no"), 
                  size = new_obs,
                  prob = c(0.7, 0.3), 
                  replace = TRUE)
)

num_pos <- my_data %>%
  filter(status == "yes") %>%
  nrow()

print(paste0("Of the ", new_obs, 
             " observations, there were ", 
             num_pos, " positive responses."))
```

Use the `plot_beta_binomial()` and `summarize_beta_binomial()` functions in the code cell below to update your <font color="purple">weakly informative `beta(6, 4)` prior with these `r new_obs` observations including `r new_pos` positive outcomes.</font>

```{webr-r}

```

:::{.callout-caution}
## Reflection Question

Visually and numerically, what insights do the results from these two function calls give you about the most likely values of your `<font color="purple">population parameter</font>?
:::

### Credible Intervals

### Highest Density Probability Intervals

## Summary

## References

:::{.refs}

:::